Northern Pikeminnow Predation on Juvenile Chinook Salmon in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Populations

### 2021-12-22 analysis/paper/paper.Rmd
draft Methods complete

The Deadwater Slough is an approximately 1.5-kilometer section of the mainstem Salmon River located roughly 5.8 river kilometers downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event that deposited substantial amounts of sediment at the confluence of the Salmon River [@Emerson1973], thereby creating a slow, deep section in the river, spanning approximately 30 acres, and averaging 68 m width.

@@

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the juvenile DSR and NRR emigrations. Our intent was to sample during the emigration of the 2018 brood year (fall 2019 and spring 2020). However, due to logistical concerns during the onset of the Covid-19 pandemic, the spring 2020 effort was delayed until fall 2020. Although this provided the benefit of an additional abundance estimate during the DSR emigration, estimates during the NRR emigration were still lacking, and so an additional effort was added in spring 2021. Fall efforts included sampling across a two-week span; however, due to budget constraints the spring 2021 effort was conducted over a single week. Fall efforts were constrained within a two-week window to minimize predator emigration/immigration and to constrain the population for the mark-recapture models.

@@

Multiple capture methods were employed during the fall 2019 effort to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. However, ineffective methods were abandoned in subsequent years. <!-- For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. Significant amounts of debris in the water column also created concerns that fyke nets could become entangled. Additionally, there were concerns that raft electrofishers and fyke nets could incidentally capture ESA-listed adult steelhead that are present in the mainstem Salmon River during fall and spring. -->After evaluating all methods, angling proved to be the most effective method for capturing piscine predators while minimizing potential impacts to ESA-listed steelhead. <!--, but recognizing that angling can have both species and size selectivity (bias).--> Therefore, the following analyses will focus on fish captured by angling, unless otherwise noted. During the fall 2019 effort it additionally became clear that Northern Pikeminnow were by far the most prevalent piscine predator in Deadwater Slough and so is the focus of our study.

@@

Four different mark-recapture abundance estimators were employed to explore variation in estimates among them. Estimators fell into two broad categories: single-census and multiple-census. For the single census estimators, we treated the first week of sampling as the mark event, and the following (second) week as the recapture week, pooling data within each of those weeks. Alternatively, the multiple census estimators treat each day as a sampling occassion, and use information about the total marked fish from all previous occassions to infer the total abundance.

For the single-census estimators, we first used the Lincoln-Peterson estimator:

$$
\hat{N} = \frac{(M)(n)}{(m)}
$$
where $M$ is the number is the total number of fish marked and returned to the population during the first week of sampling (mark event), $n$ is the total number of fish caught in the second week (recapture event), and $m$ is the number of marked fish caught during the recapture event. Because the Lincoln-Peterson estimator can be biased with small samples, we also employed the Chapman-modified Lincoln-Peterson estimator:

$$
\hat{N} = \frac{(M + 1)(n + 1)}{(m + 1)} - 1
$$
For a multiple-census estimator, we first utilized the Schnabel method:

@@

where $M$, $n$, and $m$ are indexed by the sampling occassion, $i$. Here, $i$ is defined as each sampling day. The Schnabel estimator does not have an associated standard error; however, the 90% confidence intervals were calculated. Lastly, we explored one additional multiple-census estimator, the Schumacher-Eschmeyer, which is based on minimizing the weighted sum of squares between the proportion of marked individuals in the sample and the unknown proportion of marked individuals in the population. The Schumacher-Eschmeyer estimator is as follows:

@@

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and fall 2020 sampling periods. However, to estimate the abundance of Northern Pikeminnow during spring 2021, which was not a mark-recapture design, we used the ratio of total CPUE in the fall sampling events to the total CPUE in the spring sampling event, then multiplied that ratio by the average abundance from the fall sampling events.<!-- Kevin, is this previous statement accurate? --> This approach assumes equal capture probabilities between the fall and spring capture events.

All mark-recapture abundance estimators assumed: (1) the population is closed (no immigration, emigration, births or deaths during the sampling period), (2) all fish have equal chance of being caught in the second (and subsequent) sample(s), (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, (5) marks are not missed or mistaken.

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes (e.g., juvenile steelhead, juvenile Sockeye Salmon, Redside Shiner *Richardsonius balteatus*, etc.). Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks for later analysis in a controlled environment. For each sample, wet weight (grams) was recorded for the total combined stomach content, including all non-fish items (e.g., macroinvertebrates, organic matter), as well as the portion consisting of fish parts/matter. Fish and fish remnants were identified to the lowest taxonomic unit using diagnostic bones, or were categorized as unknown [@Hansel1988; @Frost2000]. A subset of Northern Pikeminnow captures (~ 5%) were euthanized for dissection after gastric lavage to validate the efficacy of the methodology.

@@

To estimate the total consumption potential (i.e., the number or amount of juvenile Chinook Salmon consumed during the peaks of DSR [fall] and NRR [spring] emigrations) of Northern Pikeminnow in Deadwater Slough, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 and applied in the R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on the following inputs: predator and prey energy densities, predator start and end weights, and water temperatures. Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 21,500 J/g for juvenile Chinook Salmon [@Moss2016]. We used a weight-length formula [@Parker1995] to calculate the predator start weight using the average length of Northern Pikeminnow captured in Deadwater Slough during our study. The average length of Northern Pikeminnow during the fall and spring were 352.9 and 393.7 mm, respectively, which calculated to average starting weights of 504.4 and 670.5 g. The assumption of zero growth was included to estimate consumption if individual Northern Pikeminnow were not growing and/or the population is stable, and so start and end weights are equal. Mean daily water temperatures were summarized from 15-minute interval temperature readings available March 3, 2013 to June 14, 2021 from a gage station approximately 22 river kilometers downstream of Deadwater Slough near Shoup, Idaho [@USGS2016].

Our largest uncertainty for model parameters was the proportion of the Northern Pikeminnow diet consisting of fishes (e.g., juvenile Chinook Salmon) versus other food items like invertebrates. Therefore, we varied the proportion of the diet that was fish or invertebrates ranging from 10-90%, in 10% intervals, for both categories. Energy densities for invertebrates was fixed at 3,000 J/g. Separate models were also run for a 77-day fall period from September 15 - November 30 and a 92-day period in the spring from March 1 - May 31 to coincide with peak emigrations of DSR and NRR juveniles from the Lemhi River, the largest Chinook Salmon population in the Upper Salmon MPG. This resulted in 18 separate model runs (2 seasons and 9 diet scenarios). Each model run provided an estimate of the amount (grams) of fish consumed by an individual Northern Pikeminnow during the fall or spring time periods.

@@

To estimate the total biomass of fish potentially consumed by Northern Pikeminnow in Deadwater Slough, we then multiplied the biomass consumed by an individual Northern Pikeminnow for the timer period by the estimated population size for that time period (mean Schnabel estimates in the fall, CPUE ratio generated estimate in the spring). The total biomass was further converted to an estimate of the total number of Chinook salmon consumed by dividing by a weight of `r avg_DSR_g` for the fall and `r avg_NRR_g` for the spring, an average from juvenile Chinook Salmon captured during those date ranges at seven rotary screw traps located throughout the Upper Salmon upstream of Deadwater Slough.

>>>>

The Deadwater Slough is an approximately 1.5 km section of the mainstem Salmon River located approximately 6 river km downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event that deposited substantial amounts of sediment at the confluence of the Salmon River [@Emerson1973], thereby creating a slow, deep section in the river, spanning approximately 30 acres, and averaging 68 m width.

@@

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the juvenile DSR and NRR emigrations. Our initial intent was to sample during the emigration of the 2018 brood year (fall 2019 and spring 2020). However, due to logistical concerns during the onset of the Covid-19 pandemic, the spring 2020 effort was delayed until fall 2020. Although this provided the benefit of an additional abundance estimate during the DSR emigration, estimates during the NRR emigration were still lacking, and so an additional effort was added in spring 2021. Fall efforts included sampling across a two-week span, but due to budget constraints the spring 2021 effort was limited to a single week. Fall efforts were constrained within a two-week window to minimize predator emigration/immigration and to constrain the population for the mark-recapture models.

@@

Multiple capture methods were employed during the initial fall 2019 effort to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. Ineffective methods were then abandoned in subsequent years. <!-- For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. Significant amounts of debris in the water column also created concerns that fyke nets could become entangled. Additionally, there were concerns that raft electrofishers and fyke nets could incidentally capture ESA-listed adult steelhead that are present in the mainstem Salmon River during fall and spring. -->After evaluating all methods, angling proved to be the most effective method for capturing piscine predators while also minimizing potential impacts to ESA-listed adult steelhead present in the Salmon River during our sampling efforts.<!--, but recognizing that angling can have both species and size selectivity (bias).--> Therefore, the following analyses will focus on fish captured by angling, unless otherwise noted. During the fall 2019 effort it additionally became clear that Northern Pikeminnow were by far the most prevalent piscine predator in Deadwater Slough and so is the focus of our study.

@@

Four different mark-recapture estimators were employed to explore variation in abundance estimates among them. Estimators fell into two broad categories: single census and multiple census. For the single census estimators, we treated the first week as the first (mark) sample, and the following (second) week as the second (recapture) sample, pooling data within each of those weeks. Alternatively, the multiple census estimators treat each day as a sample and use information about the total marked fish from all previous samples to infer the total abundance.

For the single census estimators, we first used the Lincoln-Peterson estimator:

$$
\hat{N} = \frac{(M)(n)}{(m)}
$$
where $M$ is the total number of fish marked and returned to the population during the first sample (week), $n$ is the total number of fish caught in the second sample (week), and $m$ is the number of marked fish caught during the second week. Because the Lincoln-Peterson estimator can be biased with small samples, we also employed the Chapman-modified Lincoln-Peterson estimator:

$$
\hat{N} = \frac{(M + 1)(n + 1)}{(m + 1)} - 1
$$
For a multiple census estimator, we first utilized the Schnabel method:

@@

where $M$, $n$, and $m$ are indexed by each sample, $i$. Here, sample $i$ is defined as each day. The Schnabel estimator does not have an associated standard error; however, the 95% confidence intervals were calculated. Lastly, we explored one additional multiple-census estimator, the Schumacher-Eschmeyer, which is based on minimizing the weighted sum of squares between the proportion of marked individuals in the sample and the unknown proportion of marked individuals in the population. The Schumacher-Eschmeyer estimator is as follows:

@@

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and 2020 sample efforts (the estimate is of the initial population size). However, to estimate the abundance of Northern Pikeminnow during spring 2021, which was not a mark-recapture design, we used the ratio of total CPUE in the fall efforts to the total CPUE in the spring efforts, then multiplied that ratio by the average abundance from the fall efforts. This approach assumes equal capture probabilities between fall and spring efforts.

All mark-recapture abundance estimators assumed: (1) the population is closed (no immigration, emigration, births or deaths during the sampling period), (2) all fish have equal chance of being caught in the second (and subsequent) sample(s), (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, and (5) marks are not missed or mistaken.

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes (e.g., juvenile steelhead, juvenile Sockeye Salmon, Redside Shiner *Richardsonius balteatus*, etc.). Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks for later analysis in a controlled environment. For each sample, wet weight (grams) was recorded for the total combined stomach content, including all non-fish items (e.g., macroinvertebrates, organic matter), as well as that portion consisting of fish parts or matter. Fish and fish remnants were identified to the lowest taxonomic unit using diagnostic bones, or were categorized as unknown [@Hansel1988; @Frost2000]. A subset of Northern Pikeminnow captures (~ 5%) were euthanized for dissection after gastric lavage to validate the efficacy of the methodology.

@@

To estimate the total consumption potential (i.e., the number or amount of juvenile Chinook Salmon consumed during the peaks of DSR [fall] and NRR [spring] emigrations) of Northern Pikeminnow in Deadwater Slough, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 and applied in the R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on the following inputs: predator and prey energy densities, predator start and end weights, and water temperatures. Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 21,500 J/g for juvenile Chinook Salmon [@Moss2016]. The predator start weight was calculated using a weight-length formula [@Parker1995] and the average length of Northern Pikeminnow caught in Deadwater Slough during our study. The average length of Northern Pikeminnow caught during the fall and spring efforts were 352.9 and 393.7 mm, respectively, which calculated to average starting weights of 504.4 and 670.5 g. We assumed no growth in individual Northern Pikeminnow, and so predator start and end weights are equal, i.e., we assumed the population biomass was stable. Mean daily water temperatures were summarized from 15-minute interval temperature readings available March 3, 2013 to June 14, 2021 from a gage station approximately 22 river kilometers downstream of Deadwater Slough near Shoup, Idaho [@USGS2016].

The largest uncertainty among model parameters was the proportion of the Northern Pikeminnow diet consisting of fishes (e.g., juvenile Chinook Salmon) versus other food items (e.g., invertebrates). Therefore, we varied the proportion of the diet consisting of fish versus invertebrates to range from 10-90%, in 10% intervals, for both categories. Energy densities for invertebrates was fixed at 3,000 J/g. Separate models were also run for a 77-day fall period from September 15 - November 30 and a 92-day period in the spring from March 1 - May 31 to coincide with peak emigrations of DSR and NRR juveniles from the Lemhi River, the largest Chinook Salmon population in the Upper Salmon MPG. This resulted in 18 separate model runs (9 diet scenarios and 2 seasons). Each model run provided an estimate of the amount (grams) of fish consumed by an individual Northern Pikeminnow during the fall or spring time periods.

@@

To estimate the total biomass of DSR and NRR emigrants potentially consumed by all Northern Pikeminnow in Deadwater Slough, we then multiplied the grams of fish consumed by an individual Northern Pikeminnow  during each period (fall and spring) by the estimated population size for that period. The total biomass was further converted to an estimate of the total number of Chinook Salmon consumed each period by dividing by the average weight of DSR (`r avg_DSR_g` g) and NRR (`r avg_NRR_g` g) emigrants in the Upper Salmon MPG. The weight of an average DSR and NRR emigrants was taken from fish caught at seven rotary screw traps located throughout the Upper Salmon MPG upstream of Deadwater Slough during the fall and spring time periods.
